---
title: "cleaning data for project"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(readr)
library(lubridate)
library(dplyr)
#load data
#boston kenmore
PM25_Boston_2018 <- read_csv("project/PM25_Boston_2018.csv")
PM25_Boston_2019 <- read_csv("project/PM25_Boston_2019.csv")
PM25_Boston_2020 <- read_csv("project/PM25_Boston_2020.csv")
PM25_Boston_2021 <- read_csv("project/PM25_Boston_2021.csv")
PM25_Boston_2022 <- read_csv("project/PM25_Boston_2022.csv")
#springfield
PM25_Springfield_2018 <- read_csv("project/PM25_Springfield_2018.csv")
PM25_Springfield_2019 <- read_csv("project/PM25_Springfield_2019.csv")
PM25_Springfield_2020 <- read_csv("project/PM25_Springfield_2020.csv")
PM25_Springfield_2021 <- read_csv("project/PM25_Springfield_2021.csv")
PM25_Springfield_2022 <- read_csv("project/PM25_Springfield_2022.csv")
#worcester
PM25_Worcester_2018 <- read_csv("project/PM25_Worcester_2018.csv")
PM25_Worcester_2019 <- read_csv("project/PM25_Worcester_2019.csv")
PM25_Worcester_2020 <- read_csv("project/PM25_Worcester_2020.csv")
PM25_Worcester_2021 <- read_csv("project/PM25_Worcester_2021.csv")
PM25_Worcester_2022 <- read_csv("project/PM25_Worcester_2022.csv")
#brockton
PM25_Brockton_2018 <- read_csv("project/PM25_Brockton_2018.csv")
PM25_Brockton_2019 <- read_csv("project/PM25_Brockton_2019.csv")
PM25_Brockton_2020 <- read_csv("project/PM25_Brockton_2020.csv")
PM25_Brockton_2021 <- read_csv("project/PM25_Brockton_2021.csv")
PM25_Brockton_2022 <- read_csv("project/PM25_Brockton_2022.csv")
#clelmsford (near lowell)
PM25_Clemsford_2018 <- read_csv("project/PM25_Clelmsford_2018.csv")
PM25_Clemsford_2019 <- read_csv("project/PM25_Clemsford_2019.csv")
PM25_Clemsford_2020 <- read_csv("project/PM25_Clemsford_2020.csv")
PM25_Clemsford_2021 <- read_csv("project/PM25_Clemsford_2021.csv")
PM25_Clemsford_2022 <- read_csv("project/PM25_Clelmsford_2022.csv")
#north adams
PM25_NorthAdams_2018 <- read_csv("project/PM25_NorthAdams_2018.csv")
PM25_NorthAdams_2019 <- read_csv("project/PM25_NorthAdams_2019.csv")
PM25_NorthAdams_2020 <- read_csv("project/PM25_NorthAdams_2020.csv")
PM25_NorthAdams_2021 <- read_csv("project/PM25_NorthAdams_2021.csv")
PM25_NorthAdams_2022 <- read_csv("project/PM25_NorthAdams_2022.csv")
```

```{r}
# filter for only one Parameter Occurrence Code POC (unique monitor at the site)
# select only necessary variables
# And and rename Daily PM2.5 Concentration and POC to be unique for each city
# measurements every 3 days for entire year

PM25_Bos_2018_1 <- PM25_Boston_2018 |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Boston Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)
PM25_Bos_2019_1<- PM25_Boston_2019 |> filter(POC == 1) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Boston Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)
PM25_Bos_2020_1<- PM25_Boston_2020 |> filter(POC == 1) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Boston Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`) 
PM25_Bos_2021_1<- PM25_Boston_2021 |> filter(POC == 1) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Boston Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)
PM25_Bos_2022_1<- PM25_Boston_2022 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Boston Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)

# note that for pm2.5 boston 2021, POC = 3 and 23 have full year daily data

# pm2.5 springfield POC = 3 has full year daily data for all years
PM25_Spr_2019_1 <- PM25_Springfield_2019 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Springfield Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)
PM25_Spr_2020_1 <- PM25_Springfield_2020 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Springfield Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)
PM25_Spr_2021_1 <- PM25_Springfield_2021 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Springfield Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)

# worcester, POC=3 has full year daily data

PM25_Wor_2018_1 <- PM25_Worcester_2018 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Worcester Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)
PM25_Wor_2019_1 <- PM25_Worcester_2019 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Worcester Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)
PM25_Wor_2020_1 <- PM25_Worcester_2020 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Worcester Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)
PM25_Wor_2021_1 <- PM25_Worcester_2021 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Worcester Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)
PM25_Wor_2022_1 <- PM25_Worcester_2022 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Worcester Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)

# brockton, POC=3 has full year daily data
PM25_Bro_2018_1 <- PM25_Brockton_2018 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Brockton Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)
PM25_Bro_2019_1 <- PM25_Brockton_2019 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Brockton Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)
PM25_Bro_2020_1 <- PM25_Brockton_2020 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Brockton Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)
PM25_Bro_2021_1 <- PM25_Brockton_2021 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Brockton Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)
PM25_Bro_2022_1 <- PM25_Brockton_2022 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Brockton Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)

# clemsford, POC=3 has full year daily data
PM25_Cle_2018_1 <- PM25_Clemsford_2018 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Clemsford Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)
PM25_Cle_2019_1 <- PM25_Clemsford_2019 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Clemsford Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)
PM25_Cle_2020_1 <- PM25_Clemsford_2020 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Clemsford Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)
PM25_Cle_2021_1 <- PM25_Clemsford_2021 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Clemsford Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)
PM25_Cle_2022_1 <- PM25_Clemsford_2022 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`Clemsford Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)

# north adams, POC=3 has full year daily data
PM25_Nor_2018_1 <- PM25_NorthAdams_2018 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`North Adams Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)
PM25_Nor_2019_1 <- PM25_NorthAdams_2019 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`North Adams Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)
PM25_Nor_2020_1 <- PM25_NorthAdams_2020 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`North Adams Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)
PM25_Nor_2021_1 <- PM25_NorthAdams_2021 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`North Adams Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)
PM25_Nor_2022_1 <- PM25_NorthAdams_2022 |> filter(POC == 3) |> select(Date, `Daily Mean PM2.5 Concentration`) |> rename(`North Adams Daily Mean PM2.5 Concentration` = `Daily Mean PM2.5 Concentration`)
```

```{r}
# merge cities by date onto yearly df
#excluding springfield now because it's missing march 2020 dates
#ALSO excluding clelmsford because its missing first half of 2018
all_2018 <- plyr::join_all(list(PM25_Bos_2018_1,PM25_Wor_2018_1, PM25_Bro_2018_1, PM25_Nor_2018_1), by='Date', type='left')
all_2019 <- plyr::join_all(list(PM25_Bos_2019_1,PM25_Wor_2019_1, PM25_Bro_2019_1, PM25_Nor_2019_1), by='Date', type='left')
all_2020 <- plyr::join_all(list(PM25_Bos_2020_1,PM25_Wor_2020_1, PM25_Bro_2020_1, PM25_Nor_2020_1), by='Date', type='left')
all_2021 <- plyr::join_all(list(PM25_Bos_2021_1,PM25_Wor_2021_1, PM25_Bro_2021_1, PM25_Nor_2021_1), by='Date', type='left')
all_2022 <- plyr::join_all(list(PM25_Bos_2022_1,PM25_Wor_2022_1, PM25_Bro_2022_1, PM25_Nor_2022_1), by='Date', type='left')

# merge all into one dataset
all_PM25 <- rbind(all_2018, all_2019, all_2020, all_2021, all_2022)

# format $Date field
all_PM25$Date <- as.Date(all_PM25$Date, format='%m/%d/%Y')

# create new column for year
all_PM25$Year <- as.character(lubridate::year(all_PM25$Date))

# assign consecutive number values to each date
all_PM25$Date_Number = as.integer(as.factor(all_PM25$Date))
```

## plots, visuals
```{r}
library(astsa)
# On March 23, Governor Baker announced a stay-at-home advisory effective from noon March 24 until noon April 7
# march 10 to may 29 I think is when restrictions were in place
# this would be equal to time points 233 and 258
par(mfrow=c(2,2))
plot.ts(all_PM25$`Boston Daily Mean PM2.5 Concentration`, main="Boston", ylab="Daily Mean PM2.5 Concentration", ylim=c(0, 27))
abline(v=233, col="red", lw=1)
abline(v=258, col="red", lw=1)
# plot.ts(all_PM25$`Springfield Daily Mean PM2.5 Concentration`, main="Springfield", ylab="Daily Mean PM2.5 Concentration", ylim=c(0, 30))
# abline(v=138, col="red", lw=1)
# abline(v=163, col="red", lw=1)
plot.ts(all_PM25$`Worcester Daily Mean PM2.5 Concentration`, main="Worcester", ylab="Daily Mean PM2.5 Concentration", ylim=c(0, 27))
abline(v=233, col="red", lw=1)
abline(v=258, col="red", lw=1)
plot.ts(all_PM25$`Brockton Daily Mean PM2.5 Concentration`, main="Brockton", ylab="Daily Mean PM2.5 Concentration", ylim=c(0, 27))
abline(v=233, col="red", lw=1)
abline(v=258, col="red", lw=1)
# plot.ts(all_PM25$`Clemsford Daily Mean PM2.5 Concentration`, main="Clemsford", ylab="Daily Mean PM2.5 Concentration", ylim=c(0, 30))
# abline(v=138, col="red", lw=1)
# abline(v=163, col="red", lw=1)
plot.ts(all_PM25$`North Adams Daily Mean PM2.5 Concentration`, main="North Adams", ylab="Daily Mean PM2.5 Concentration", ylim=c(0, 27))
abline(v=233, col="red", lw=1)
abline(v=258, col="red", lw=1)

```

```{r}
par(mfrow=c(2, 2))
plot.ts(PM25_Bos_2018_1$`Boston Daily Mean PM2.5 Concentration`, main="Boston", ylab="Daily Mean PM2.5 Concentration", ylim=c(0, 25), col="pink", lw=1.2)
lines(PM25_Bos_2019_1$`Boston Daily Mean PM2.5 Concentration`, col="orchid", lw=1.2)
lines(PM25_Bos_2020_1$`Boston Daily Mean PM2.5 Concentration`, col="black", lw=2)
lines(PM25_Bos_2021_1$`Boston Daily Mean PM2.5 Concentration`, col="lightskyblue", lw=1.2)
lines(PM25_Bos_2022_1$`Boston Daily Mean PM2.5 Concentration`, col="seagreen3", lw=1.2)
abline(v=22, col="black", lw=1, lty=2)
abline(v=49, col="black", lw=1, lty=2)

legend('topright', legend = c('2018', '2019', '2020', '2021', '2022'), col = c('pink', 'orchid', 'black', 'lightskyblue', 'seagreen3'), lty = 1, lwd=2)

# use same 3 day pattern as in all_PM25 data
PM25_Wor_2018_2 <- all_PM25 |> select(Date, `Worcester Daily Mean PM2.5 Concentration`, Year) |>filter(Year==2018)
PM25_Wor_2019_2 <- all_PM25 |> select(Date, `Worcester Daily Mean PM2.5 Concentration`, Year) |>filter(Year==2019)
PM25_Wor_2020_2 <- all_PM25 |> select(Date, `Worcester Daily Mean PM2.5 Concentration`, Year) |>filter(Year==2020)
PM25_Wor_2021_2 <- all_PM25 |> select(Date, `Worcester Daily Mean PM2.5 Concentration`, Year) |>filter(Year==2021)
PM25_Wor_2022_2 <- all_PM25 |> select(Date, `Worcester Daily Mean PM2.5 Concentration`, Year) |>filter(Year==2022)

plot.ts(PM25_Wor_2018_2$`Worcester Daily Mean PM2.5 Concentration`, main="Worcester", ylab="Daily Mean PM2.5 Concentration", ylim=c(0, 25), col="pink", lw=1.2)
lines(PM25_Wor_2019_2$`Worcester Daily Mean PM2.5 Concentration`, col="orchid", lw=1.2)
lines(PM25_Wor_2020_2$`Worcester Daily Mean PM2.5 Concentration`, col="black", lw=2)
lines(PM25_Wor_2021_2$`Worcester Daily Mean PM2.5 Concentration`, col="lightskyblue", lw=1.2)
lines(PM25_Wor_2022_2$`Worcester Daily Mean PM2.5 Concentration`, col="seagreen3", lw=1.2)
abline(v=22, col="black", lw=1, lty=2)
abline(v=49, col="black", lw=1, lty=2)

# use same 3 day pattern as in all_PM25 data
PM25_Bro_2018_2 <- all_PM25 |> select(Date, `Brockton Daily Mean PM2.5 Concentration`, Year) |>filter(Year==2018)
PM25_Bro_2019_2 <- all_PM25 |> select(Date, `Brockton Daily Mean PM2.5 Concentration`, Year) |>filter(Year==2019)
PM25_Bro_2020_2 <- all_PM25 |> select(Date, `Brockton Daily Mean PM2.5 Concentration`, Year) |>filter(Year==2020)
PM25_Bro_2021_2 <- all_PM25 |> select(Date, `Brockton Daily Mean PM2.5 Concentration`, Year) |>filter(Year==2021)
PM25_Bro_2022_2 <- all_PM25 |> select(Date, `Brockton Daily Mean PM2.5 Concentration`, Year) |>filter(Year==2022)

plot.ts(PM25_Bro_2018_2$`Brockton Daily Mean PM2.5 Concentration`, main="Brockton", ylab="Daily Mean PM2.5 Concentration", ylim=c(0, 25), col="pink", lw=1.2)
lines(PM25_Bro_2019_2$`Brockton Daily Mean PM2.5 Concentration`, col="orchid", lw=1.2)
lines(PM25_Bro_2020_2$`Brockton Daily Mean PM2.5 Concentration`, col="black", lw=2)
lines(PM25_Bro_2021_2$`Brockton Daily Mean PM2.5 Concentration`, col="lightskyblue", lw=1.2)
lines(PM25_Bro_2022_2$`Brockton Daily Mean PM2.5 Concentration`, col="seagreen3", lw=1.2)
abline(v=22, col="black", lw=1, lty=2)
abline(v=49, col="black", lw=1, lty=2)

# use same 3 day pattern as in all_PM25 data
PM25_Nor_2018_2 <- all_PM25 |> select(Date, `North Adams Daily Mean PM2.5 Concentration`, Year) |>filter(Year==2018)
PM25_Nor_2019_2 <- all_PM25 |> select(Date, `North Adams Daily Mean PM2.5 Concentration`, Year) |>filter(Year==2019)
PM25_Nor_2020_2 <- all_PM25 |> select(Date, `North Adams Daily Mean PM2.5 Concentration`, Year) |>filter(Year==2020)
PM25_Nor_2021_2 <- all_PM25 |> select(Date, `North Adams Daily Mean PM2.5 Concentration`, Year) |>filter(Year==2021)
PM25_Nor_2022_2 <- all_PM25 |> select(Date, `North Adams Daily Mean PM2.5 Concentration`, Year) |>filter(Year==2022)

plot.ts(PM25_Nor_2018_2$`North Adams Daily Mean PM2.5 Concentration`, main="North Adams", ylab="Daily Mean PM2.5 Concentration", ylim=c(0, 25), col="pink", lw=1.2)
lines(PM25_Nor_2019_2$`North Adams Daily Mean PM2.5 Concentration`, col="orchid", lw=1.2)
lines(PM25_Nor_2020_2$`North Adams Daily Mean PM2.5 Concentration`, col="black", lw=2)
lines(PM25_Nor_2021_2$`North Adams Daily Mean PM2.5 Concentration`, col="lightskyblue", lw=1.2)
lines(PM25_Nor_2022_2$`North Adams Daily Mean PM2.5 Concentration`, col="seagreen3", lw=1.2)
abline(v=22, col="black", lw=1, lty=2)
abline(v=49, col="black", lw=1, lty=2)
```





## trying smoothing methods:

Checking for stationarity, overall trends:

```{r}
Nor_ts <- ts(all_PM25$`North Adams Daily Mean PM2.5 Concentration`)
acf(north_adams_ts)
```

NOTE: ACF cannot be run on data with missing values

smoothing methods to identify seasonal trends:

1.) Moving Average Smoothers

```{r}
plot(Nor_ts, type = 'o', col = 'black', 
     xlab = 'Time', ylab = 'pm2.5', main = 'North Adams: Moving Average Smoothers: k=2, k=5')
wgts1 = c(1/8, rep(1/4, 3), 1/8)
lhf1 = stats::filter(Nor_ts, sides=2, filter=wgts1)
lines(lhf1, lwd=2, col="blue")
wgts2 = c(1/20, rep(1/10, 9), 1/20)
lhf2 = stats::filter(Nor_ts, sides=2, filter=wgts2)
lines(lhf2, lwd=2, col="red")
legend('topleft', legend = c('k=2', 'k=5'), col = c('blue', 'red'), lty = 1, lwd=2)

abline(v=138, col="red", lw=1)
abline(v=163, col="red", lw=1)


```

2.) LOESS - k nearest neighbors 2 & 8 "For 2 nearest neighbors, the f value (percent of points in dataset that are considered nearest neighbors) would be $f=2/n$ and for 8 nearest neighbors, the f value would be $f=8/n$, since the n value here is 333 observation. Then we are applying LOWESS for $f=0.006006$ and $f=0.024024$"

```{r}
plot(Nor_ts, type = 'o', col = 'black', 
     xlab = 'Time', ylab = 'LakeHuron', main = 'LOWESS with 2 and 8 neighbors')
# wgts1 = c(1/8, rep(1/4, 3), 1/8)
# lhf1 = stats::filter(Nor_ts, sides=2, filter=wgts1)
# lines(lhf1, lwd=2, col="blue")
# wgts2 = c(1/20, rep(1/10, 9), 1/20)
# lhf2 = stats::filter(Nor_ts, sides=2, filter=wgts2)
# lines(lhf2, lwd=2, col="red")
# legend('topleft', legend = c('k=2', 'k=5'), col = c('blue', 'red'), lty = 1, lwd=2)
lines(lowess(Nor_ts, f=2/333), lwd=1.5, col="orange") 
lines(lowess(Nor_ts, f=8/333), lwd=1.5, col="green")
legend('topright', legend = c('f=2/n', 'f=8/n'), col = c('orange', 'green'), lty = 1, lwd=2)
```
